Gelation as a condensation frustrated by hydrodynamics and mechanical tension

### 2012-08-07 paper.tex
Add binodal line

We use a novel experimental set-up that allows the observation of the entire process of gelation at a single particle level (see Methods). In a suspension of sticky colloids with long range repulsion, we suddenly turn off the repulsion. The short range attraction become dominant, eventually leading to the formation of a gel. The whole process takes place on the microscope stage and can thus be observed in three dimensions from the very beginning.

@@

Our protocol is not equivalent to a redispersion of an existing gel at the same state point either by shaking or shear melting like in  e.g.~\cite{lu2008gelation,Teece2011,Bartlett2012}. It is closer to a quench and to application relevant processes where the sudden addition of a component (e.g. salt, acid, depletion agent) induces aggregation in an otherwise stable suspension (e.g. milk).

@@

The data were collected on a Leica SP5 confocal microscope, using \unit{532}{\nano\meter} laser excitation. The temperature was controlled on both stage and objective lens, allowing a more precise density matching. The scanned volume is at least \unit{82 \times 82 \times 85}{\micro\metre}. Standard particle tracking techniques allow us to localise our particles with errors below $1\%$ of a diameter. We carefully set the scanning time in order to track more than \unit{98}\% of colloids between two successive time steps.

From direct confocal measurement~\cite{Royall2007, Poon2012}, we estimate the hard-core diameter of our colloids ($\sigma=\unit{2.75}{\micro\metre}$) and the range of the interaction potential (that confirmed our scaling of $R_g$ within $1\%$), leading to a polymer-colloid size ratio $q = 2R_g/\sigma = 0.10(6)$. Spinodal and binodal lines on Fig.~\ref{fig:methods}b are calculated from this size ratio using the generalized free volume theory~\cite{Fleer2008}. Two particles are considered bonded when their centres are within $\sigma+2R_g$, which corresponds to the first minimum of $g(r)$.

@@

\clearpage
\begin{figure}
	\begin{center}
	\includegraphics{generate-figure0.pdf}
	\end{center}
	\caption{\textbf{Observing gelation.} \textbf{a} Sketch of our experimental setup. The observation cell contains initially colloids, polymer and no salt. \textbf{b} Phase diagram of our system (the line is a guide for the eye). \textbf{c-h} Early stage of gelation observed experimentally (state point is circled in \textbf{b}). Pictures are computer reconstruction of the experimental coordinates coloured by the number of particles in clusters.}
	\label{fig:methods}
\end{figure}

>>>>

We use a novel experimental set-up that allows the observation of the entire process of gelation at a single particle level (see Methods). In a suspension of sticky colloids with long range repulsion, we suddenly turn off the repulsion. The short range attraction become dominant, eventually leading to the formation of a gel. The whole process takes place on the microscope stage and can thus be observed dynamically in three dimensions from the very beginning.

Our protocol is not equivalent to a redispersion of an existing gel at the same state point either by shaking or shear melting like in  e.g.~\cite{lu2008gelation,Teece2011,Bartlett2012}. It is closer to a quench, however we expect differences compared to a quench simulated via molecular or Brownian dynamics, because of the presence of a solvent mediating hydrodynamic interactions~\cite{Furukawa2010}. Thus our protocol might be very close to application relevant processes where the sudden addition of a component (e.g. salt, acid, depletion agent) induces aggregation in an otherwise stable suspension (e.g. milk).

@@

@@

The data were collected on a Leica SP5 confocal microscope, using \unit{532}{\nano\meter} laser excitation. The temperature was controlled on both stage and objective lens, allowing a more precise density matching. The scanned volume is at least \unit{82 \times 82 \times 85}{\micro\metre}. Standard particle tracking techniques allow us to localise our particles with errors below $1\%$ of a diameter. We carefully set the scanning time in order to be able to follow between two successive time steps all the colloids that stay in the field of view.

From direct confocal measurement~\cite{Royall2007, Poon2012}, we estimate the hard-core diameter of our colloids ($\sigma=\unit{2.75}{\micro\metre}$) and the range of the interaction potential (that confirmed our scaling of $R_g$ within $1\%$), leading to a polymer-colloid size ratio $q = 2R_g/\sigma = 0.10(6)$. Spinodal and binodal lines on Fig.~\ref{fig:methods}b are calculated from this size ratio using the generalized free volume theory~\cite{Fleer2008}. In principle two particles are considered bonded when their centres are within $\sigma+2R_g$, however resolution-dependent tracking imprecisions and systematic errors do not give a precise enough estimate of such short distance. Therefore we use the first minimum of $g(r)$ as a threshold, which is $\unit{3.40 \text{ or } 3.55}{\micro\metre}$ depending on the resolution.

@@

\clearpage
\begin{figure}
	\begin{center}
	\includegraphics{generate-figure0.pdf}
	\end{center}
	\caption{\textbf{Observing gelation.} \textbf{a} Sketch of our experimental setup. The blue arrows stands for the diffusion of the salt through the membrane into the observation cell. \textbf{b} Phase diagram of our system. Theoretical lines represent the gas-liquid spinodal (full) and binodal (dashed) as well as the gas-solid coexistence. \textbf{c-h} Early stage of gelation observed experimentally (state point is circled in \textbf{b}). Pictures are computer reconstruction of the experimental coordinates coloured by the number of particles in clusters.}
	\label{fig:methods}
\end{figure}
